Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes

### 2020-05-08 Ben_Manuscripts/hdphmm/Draft.tex
Comments on draft.

  \begin{abstract}

  Appropriate time series modeling of complex diffusion on the microsecond time scale
  can provide a path towards inferring transport mechanisms and predicting bulk properties
  characteristic of much longer time scales. In this work we apply the infinite hidden 
  Markov model (IHMM) to solute center-of-mass trajectories generated from long 
  molecular dynamics simulations in a cross-linked inverted hexagonal phase lyotropic 
  liquid crystal (LLC) membrane in order to automatically detect and parameterize an unknown 
  number of distinct latent vector autoregressive dynamical modes.

  We group similar solute behavior into clusters, analyze the dynamics characteristic
  to each cluster and relate them to physical interactions between solutes and the 
  membrane. This enhancement to our understanding of the diverse range of influential 
  solute-membrane interactions allows us hypothesize design changes to LLC monomers 
  aimed towards controlling the rates of solute passage, thus improving the selective 
  performance of LLC membranes.

@@

  Overall, this work provides an effective way to simultaneously identify transport 
  mechanisms and project complex diffusive behavior on long time scales.

@@

  % BJC: Not sure if this should be the first paragraph, or if I could just
  % delete this paragraph and start with the next one. 
  There is a need for highly selective membranes in order to perform efficient 
  separations of the components of complex aqueous streams.
  \begin{itemize}
    \item Desalination and boric acid removal from seawater.
    \item Organic micropollutants
    \item Hydraulic fracturing flowback water
    \item Breathable chemical warfare agent barriers
    \item High purity chemicals
    \item While many researchers focus on membrane permeability, we may be 
    able to reduce costs of commercial nanofiltration and reverse osmosis with
    higher selectivity.~\cite{werber_materials_2016}
  \end{itemize}

  The ability to efficiently design novel membrane materials for selective separations would be 
  greatly enhanced by easily interpretable methods for connecting microscopic fluctuations
  with macroscopic observables.
  \begin{itemize}
  	\item If one could apply molecular dynamics (MD) simulations to explain important
  	experimental metrics such as solute flux and selectivity in terms of detailed 
  	chemically dependent solute motion, then we can more easily bridge the disconnect
  	between theory and application.
  	%BJC: this could potentially be moved further down
  	\item There are two main challenges associated with this goal.
  	%since complex and
  	%diverse diffusive motion on time scales accessible to molecular simulations prevents
  	%us from easily relating chemically dependent solute motion to experimental time scales.
  	\item First, we need a simple way to identify and parameterize the fluctuations
  	characteristic to dominant solute-membrane interactions.
  	\item Second, we need a way to propagate this behavior far beyond the time scales
  	afforded by MD.
  \end{itemize}

  In this work, we are particularly interested in the design of lyotropic liquid crystal (LLC)
  monomers, a class of amphiphilic molecules whose ordered self-assembled phases can be 
  cross-linked into mechanically strong membranes capable of highly selective separations.
  \begin{itemize}
  	\item Inverted hexagonal, H\textsubscript{II}, phase LLC membranes are characterized
  	by hexagonally packed, uniform-sized and straight pores, an ideal geometry for high 
  	throughput transport.
	\item Their pores are lined with the LLC monomer functional groups which can
	potentially be designed to interact with solutes in a chemically-specific manner.
	\item It would be extremely useful to experimentalists if we could connect LLC monomer
	design with measurable quantities.
  \end{itemize}

@@

  In this work we attempt to minimize human effort by applying the infinite hidden 
  Markov Model (IHMM)~\cite{beal_infinite_2002,calderon_inferring_2015}, a nonparameteric 
  Bayesian classification technique, in order to automatically detect and infer the 
  parameters of an unknown number of latent vector autoregressive (VAR) modes present 
  in solute center-of-mass time series trajectories. 
  \begin{itemize}
    \item The IHMM has not been used much for MD trajectory analysis.
    \item It's been used extensively for speaker diarization and has found some application
    in single particle trajectory analysis.
  \end{itemize}

  We aim to address two primary scientific 
  questions with this work. 
  %BJC: this may be redundant from the second paragraph. Maybe better suited here?
  First, we want
  to know if the IHMM algorithm can help researchers efficiently uncover underlying
  transport mechanisms which give rise to different solute dynamical behavior.
  Second, we want to evaluate the ability of our model to reproduce solute trajectories
  in a stochastic manner so that we can use it to predict macroscopic transport 
  properties which would otherwise require much longer MD simulations.

@@

  Hidden Markov models (HMMs) are a useful and widely used technique for modeling
  sequences of observations where the probability of the next observation in a 
  sequence depends, at least in part, on a previous unobserved, latent or hidden,
  state.~\cite{beal_infinite_2002} In the context of our simulations, the observations
  correspond to the center of mass coordinates of the solutes versus time, and the
  states correspond to the dynamical behavior which give rise to those types
  of observations. The probability of transitioning to a state based on the current
  state is mathematically defined in terms of an $n\times n$ transition probability
  matrix, $T$, where $n$ is the number of states. Unfortunately, standard HMMs 
  require $n$ to be known \textit{a priori}. One can partially overcome this by 
  testing a range of numbers of hidden states and determining which is the best 
  representation of the data.~\cite{pohle_selecting_2017}

@@

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences. We evaluated the convergence of the state sequence based on
  the convergence of the VAR parameters. The parameters of each state
  generally converge within 500--1000 iterations. However, the boundaries of 
  each state segment fluctuate which can lead to high variance in the $A$
  and $\Sigma$ parameters in segments with a relatively low number of emissions.
  In these cases, the mean in each dimension tends to be a more reliable indicator of
  convergence (see Figure~\ref{S-fig:convergence3d} of the Supporting Information).

  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) because $x$ and $y$ aren't 
  meaningful alone due to the system's radial symmetry. We converted the $x$ and 
  $y$ center-of-mass coordinates to $r$. We fixed the state sequence from the 
  3D parameterization and applied the inference component of the IHMM procedure
  to the cylindrical trajectories in order to estimate the VAR(1) parameters in
  terms of $r$ and $z$. Since we fixed the state sequence, and we are using conjugate
  priors, the VAR parameters converge very quickly (see Figure~\ref{S-fig:fixed_state_convergence}).
  Therefore, we only ran the inference procedure for 100 iterations. 
  %However, note that the uncertainty in their
  %estimates are a function of the number of data points used in their estimate.

  We defined the finalized parameters of each state as the average of the parameters 
  from each iteration recorded after the equilibration time point. We found the 
  equilibration time point of each component of the $\mathbf{c}$ vector as well as 
  $A$ and $\Sigma$ matrices, and then used the longest equilibration time of all 
  dimensions as the equilibration time point.

@@

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size. For each solute studied, we identified 200-325
  independent states, each with separate VAR(1) parameters. Many of these states
  exhibit very similar dynamical behavior except their mean levels in $r$ and $z$
  are different.

  We reduced the parameter space used agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached.~\cite{pedregosa_scikit-learn_2011} We used the Ward linkage 
  criteria, which works to minimize the sum of the squared differences within all
  clusters.~\cite{ward_hierarchical_1963} We elected to choose the number of clusters
  rather than the distance threshold. For our data, non-parametric methods such as 
  Bayesian Gaussian mixture models~\cite{pedregosa_scikit-learn_2011,gelman_bayesian_2013}
  tend to delocalize the clusters in parameter space (see Section~\ref{S-section:agglomerative}
  of the Supporting Information).

  We clustered based on the eigenvalues of $A$ and $\Sigma$, the radial means, $\mu_r$, 
  and the self-transition probabilities, $T_{ii}$, of each state for a total of 6 clustering
  dimensions. The means in $r$ are likely meaningful since the membrane is inhomogeneous, 
  transitioning from hydrophilic pores to hydrophobic tails. To cluster on $T_{ii}$, we first
  cast it in terms of the expected dwell times. The expected value of the number of sequential
  self-transitions is simply $\frac{1}{1 - T_{ii}}$. This relationship correctly implies that 
  dwell times approach infinity as $T_{ii}$ approaches 1. There we found the most success with
  agglomerative clustering by linearizing this relationship and clustering on $-log(1 - T_{ii})$
  (see Section\ref{S-section:selfT_cluster} of the Supporting Information for a graphical
  illustration).

@@

  %BJC: to be reworked after I have a working procedure.
  Choosing the number of clusters can be somewhat subjective so
  we attempted to add some structure to the selection process by following
  a set of qualitative and quantitative guidelines. We used the silhouette 
  test in order to score the quality of clustering as a function of the number 
  of clusters chosen.~\cite{kaufman_finding_2009} For our data, the silhouette test generally favors the 
  lowest number of clusters possible (see Section~\ref{S-section:nclusters} of
  the Supporting Information). However, choosing too few clusters tends to 
  not distinguish between visually obvious differences in 
  dynamic behavior. This results in finalized parameter sets that are averages of
  distinct behavior which presents further problems with the predictive modeling
  that we discuss later on. We aimed to maintain the highest silhouette score,
  and thus lowest number of clusters, possible while verifying that visually 
  distinct states stayed separated. Based on these guidelines, we decided to 
  group the $A$ and $\Sigma$ parameters into five clusters each and the radial
  means into three clusters. Of the 75 distinct states allowed by this formulation,
  the solutes in this study showed behavior from 32--36 distinct clusters. While
  this may seem like a high number of states, many are sampled infrequently.

@@

  \subsection{Tools for Exploring Mechanisms}\label{method:interactions}

@@

  We developed a way to measure the local density of the membrane at arbitrary
  points in the unit cell. We histogrammed the positions in three dimensions
  Since our system is in a monoclinic unit cell, we periodically replicated
  the system in the +/- x, y and z directions and then chose the bounds on the
  histogram in order to create a rectangular box encompassing the unit cell with
  a 1 nm buffer between the histogram and unit cell boundaries.~\cite{van_der_walt_numpy_2011} We then used a regular
  grid interpolator in order to allow interpolation at arbitrary points within the grid.~\cite{virtanen_scipy_2020}

@@

  Solutes in this system show a wide range of behavior influenced by the 
  heterogeneity of the membrane's nanostructure as well as the interactions 
  between monomer and solute chemical functionality. Our application of the 
  IHMM can identify and distinguish these behaviors. The following approach 
  to analysis demonstrates how one can discover and explain the very complex
  behavior exhibited by solutes over long MD trajectories in terms of 
  solute-membrane interactions. Further, we can use the IHMM in order to generate
  stochastic trajectories, with dynamics similar to MD, so that we can estimate 
  macroscopic transport properties.

@@

  \subsubsection*{The Dynamics of Methanol}

@@

  \begin{figure}
  \centering
  \begin{subfigure}{0.75\textwidth}
  \includegraphics[width=\textwidth]{mechanism_map.png}
  \caption{}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{monomer_oxygens.png}
  \caption{}
  \end{subfigure}
  \caption{(a) We can learn a significant amount of detail about solute motion by viewing 
  solute trajectories color-coded according to their clustered dynamical behavior alongside
  plots of the dominant molecular interactions and descriptions of the membrane's local 
  environment. In the plot above, we show the radial and axial coordinates of a single
  methanol molecule over the course of the equilibrated portion of its MD simulation.
  Disjoint segments of the same color imply the solute behaves similarly at different 
  points of the trajectory because their VAR parameters place them in the same cluster. 
  There are eight distinct dynamical modes exhibited by this trajectory. In order
  of their first appearance, they are colored dark yellow, red, light blue, light green,
  orange, blue, teal and lime green. On the top of the plot are bars which represent different
  physical interactions and the membrane's local environment as a function of time. The top 
  bar is colored green at time points where solutes are associated with sodium. Methanol 
  associates with sodium ions relatively infrequently. The next bar down is colored when hydrogen
  bonds exist. The colors match the enlarged oxygen atoms highlighted in (b). Blue slivers
  correspond to hydrogen bonds donated to the monomer's carboxylate head group. Green slivers
  correspond to  hydrogen bonds donated to the ether linkages between the head groups
  and tails. Red slivers correspond to hydrogen bonds donated to the oxygens on the ends
  of the monomer tails. Methanol exhibits significant hydrogen bonding with all parts of
  the monomer. Which part of the monomer methanol hydrogen bonds to is heavily correlated
  to the solute's $r$ coordinate. Finally, the bar with the blue gradient represents the local 
  membrane density based on the solute's ($r$, $z$) coordinate. Darker shades of blue
  correspond to higher membrane densities. The density appears to be higher when methanol
  motion is restricted (e.g. from 75 to 550 ns). It is relatively low in areas where methanol's
  position fluctuates significantly (e.g. from 550 to 700 ns). 
  }\label{fig:mechanism_map}
  \end{figure}

@@

  We can begin to dissect the dynamics of each state by studying their parameters.
  States 1, 3 and 5 have radial means near 1 nm or less (see y axis of $r$ direction 
  plot in Figure~\ref{fig:common_states_MET_lines}), which suggest that they occur
  close to the pore centers. States 2, 4, 6 and 7 occur well within the tail region
  of the membrane. States 1 and 2 have higher covariances in both the axial and radial
  directions compared to all other states. Covariance in each dimension appears to have
  a strong correlation. States 3 and 5 (in the pore region) as well as states 4, 6 and
  7 (in the tail region) are primarily distinguished based on their autoregressive 
  coefficients, A. Higher values of A generally lead to more wandering about the state
  mean. One could argue that the parameters of states 3--7 are actually very close which
  suggests that solutes might behave alike in topologically distinct regions of the 
  membrane.

  Despite some similarity in their VAR parameters, the length of time spent in states 
  3--7 appears to be dependent on the solute's radial location. Based on the 
  self-transition probabilities, we estimated the average time spent by each solute in
  each state (Figure~\ref{fig:dwell_times_MET}). States 3, 5 and 7 have long dwell times
  relative to the rest because they have high self-transition probabilities. Somewhat 
  surprisingly, both states 3 and 5 correspond to solutes located close to a pore center,
  where one might expect solute motion to be less restricted. Although solutes in states
  4 and 6 are in the viscous tail region, their dwell times are only intermediate. This
  could be due to inhomogeneity in the tail region which causes transitions between many
  different trapped states but with little net effect on solute MSD. However, there are
  instances of states within the tail region with dwell times that exceed 100--200 ns,
  as shown by our inclusion of state 7. There are no other states with dwell times longer
  than state 5 in the pore region.

@@

  The types of hydrogen bonding undergone by solutes is dependent on radial
  location (see Figure~\ref{fig:hbond_pichart}). States with radial means near 1 nm
  or less (states 1, 3 and 5) typically hydrogen bond with the monomer carboxylate
  groups and ether linkages. States with radial means greater than 1 nm (states 2,
  4, 6 and 7) tend to hydrogen bond with the monomer tails. In the past we've shown
  that there is a gradual transition from the hydrophilic to hydrophobic region in
  these membranes, meaning the monomer location is not tightly bound in $r$. Therefore,
  there are instances of hydrogen bond interactions that one might consider 
  uncharacteristic based on the state's radial mean.

@@

  %BJC5: something we can do for the supporting information to show that this approach is valid: Eliminate nonidealities in our MD trajectories (compared to a pure VAR process) and use toy data to generate trajectories based on some number of states with different parameters. Generate an MSD with those. See if we can reproduce with our model. Can also use this to test new strategies.
  %MRS6: toy data is actually a VERY good idea.  Use a VAR process that looks like MD, and then try to cluster it, and figure out why different.

@@

  We showcase this modeling approach by example, but it is important to
  recognize the generality of this analysis, especially in the context of molecular
  simulations. Vector autogregressive models can describe a diverse set of behavior
  and so the IHMM may be suited to study many types motion, including those not 
  characterized by hopping and trapping. The IHMM is a powerful approach for 
  understanding particle motion as well as for generating cheap models
  which can give macroscopic insight.

>>>>

  \begin{abstract}
  %MRS: added ``in soft matter systems'' to make it clear that the focus on understanding things in physical systems.
  Appropriate time series modeling of complex diffusion in soft matter systems on the microsecond time scale
  can provide a path towards inferring transport mechanisms and predicting bulk properties
  characteristic of much longer time scales. In this work we apply the infinite hidden 
  Markov model (IHMM) to solute center-of-mass trajectories generated from long 
  molecular dynamics simulations in a cross-linked inverted hexagonal phase lyotropic 
  liquid crystal (LLC) membrane in order to automatically detect and parameterize an unknown 
  number of distinct latent vector autoregressive dynamical modes. 
  %MRS: in above sentence: try to rephrase around the physical information obtained (mechanism or properties) rather than the mathematical route to it.

  %MRS: below - it's not the solute behavior, though, it's states identified by them model. Rephrase?
  We group similar solute behavior into clusters, analyze the dynamics characteristic
  to each cluster and relate them to physical interactions between solutes and the 
  membrane. This enhancement to our understanding of the diverse range of influential 
  solute-membrane interactions allows us hypothesize design changes to LLC monomers 
  aimed towards controlling the rates of solute passage, thus improving the selective 
  performance of LLC membranes.

@@

  Overall, this work provides an effective way to simultaneously identify transport 
  mechanisms 
  %MRS: emphasizing things, not just analysis.
  in porous nanoscale materials
  and project complex diffusive behavior on long time scales.

@@

  % BJC: Not sure if this should be the first paragraph, or if I could just
  % delete this paragraph and start with the next one. 
  % MRS: going to JPC, so emphasize the physical systems.
  There is a need for highly selective membranes in order to perform efficient 
  separations of the components of complex aqueous streams.
  \begin{itemize}
    \item Desalination and boric acid removal from seawater.
    \item Organic micropollutants
    \item Hydraulic fracturing flowback water
    \item Breathable chemical warfare agent barriers
    \item High purity chemicals
    \item While many researchers focus on membrane permeability, we may be 
    able to reduce costs of commercial nanofiltration and reverse osmosis with
    higher selectivity.~\cite{werber_materials_2016}
  \end{itemize}

  The ability to efficiently design novel membrane materials for selective separations would be 
  greatly enhanced by easily interpretable methods for connecting microscopic fluctuations
  with macroscopic observables.
  %MRS: should explicitly point out that just calculating diffusion coefficients don't work for subdiffusive behavior, so
  %MRS need a different way to do it.
  \begin{itemize}
  	\item If one could apply molecular dynamics (MD) simulations to explain important
  	experimental metrics such as solute flux and selectivity in terms of detailed 
  	chemically dependent solute motion, then we can more easily bridge the disconnect
  	between theory and application.
  	%BJC: this could potentially be moved further down
  	\item There are two main challenges associated with this goal.
  	%since complex and
  	%diverse diffusive motion on time scales accessible to molecular simulations prevents
  	%us from easily relating chemically dependent solute motion to experimental time scales.
  	\item First, we need a simple way to identify and parameterize the fluctuations
  	characteristic to dominant solute-membrane interactions.
  	\item Second, we need a way to propagate this behavior far beyond the time scales
  	afforded by MD.
  \end{itemize}

  %MRS: point out that LLC's are examples where there are lots of hopping/jumping (cite previous papers) and
  %MRS: so we need different approaches.
  In this work, we are particularly interested in the design of lyotropic liquid crystal (LLC)
  monomers, a class of amphiphilic molecules whose ordered self-assembled phases can be 
  cross-linked into mechanically strong membranes capable of highly selective separations.
  \begin{itemize}
  	\item Inverted hexagonal, H\textsubscript{II}, phase LLC membranes are characterized
  	by hexagonally packed, uniform-sized and straight pores, an ideal geometry for high 
  	throughput transport.
	\item Their pores are lined with the LLC monomer functional groups which can
	potentially be designed to interact with solutes in a chemically-specific manner.
	\item It would be extremely useful to experimentalists if we could connect LLC monomer
	design with measurable quantities.
  \end{itemize}

@@

  In this work we attempt to minimize human effort by applying the infinite hidden 
  Markov Model (IHMM)~\cite{beal_infinite_2002,calderon_inferring_2015}, a nonparameteric 
  Bayesian classification technique, in order to automatically detect and infer the 
  parameters of an unknown number of latent vector autoregressive (VAR) modes present 
  in solute center-of-mass time series trajectories. 
  \begin{itemize}
    %MRS: start with where it has been used before (to establish utility) a
    \item It's been used extensively for speaker diarization and has found some application
    in single particle trajectory analysis.
    %MRS: Then say it hasn't been used in this context (do we know of any, if not say so).
    \item The IHMM has not been used much for MD trajectory analysis.

  \end{itemize}

  We aim to address two primary scientific 
  questions with this work. 
  %BJC: this may be redundant from the second paragraph. Maybe better suited here?
  First, we want
  to know if the IHMM algorithm can help researchers efficiently uncover underlying
  transport mechanisms which give rise to different solute dynamical behavior.
  %MRS: flipping for clarity.
  %Second, we want to evaluate the ability of our model to reproduce solute trajectories
  %in a stochastic manner so that we can use it to predict macroscopic transport 
  %properties which would otherwise require much longer MD simulations.
  Second, we want to predict macroscopic transport by using our model to generate 
  stochastic solute trajectories at much longer timescales than possible with MD simulations, 
  extending into timescale where diffusive behavior again occurs.

@@

  %MRS: consider added subsubsections to make it more readable.
  Hidden Markov models (HMMs) are a useful and widely used technique for modeling
  sequences of observations where the probability of the next observation in a 
  sequence depends, at least in part, on a previous unobserved, latent or hidden,
  state.~\cite{beal_infinite_2002} In the context of our simulations, the observations
  correspond to the center of mass coordinates of the solutes versus time, and the
  states correspond to the dynamical behavior which give rise to those types
  of observations. The probability of transitioning to a state based on the current
  state is mathematically defined in terms of an $n\times n$ transition probability
  matrix, $T$, where $n$ is the number of states. Unfortunately, standard HMMs 
  require $n$ to be known \textit{a priori}. One can partially overcome this by 
  testing a range of numbers of hidden states and determining which is the best 
  representation of the data.~\cite{pohle_selecting_2017}

@@

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences. We evaluated the convergence of the state sequence based on
  the convergence of the VAR parameters. The parameters of each state
  generally converge within 500--1000 iterations. However, the boundaries of 
  each state segment fluctuate which can lead to high variance in the $A$
  and $\Sigma$ parameters in segments with a relatively low number of emissions.
  In these cases, 
  %MRS: maybe plateauing of the mean below?  Be more precise.
  the mean in each dimension tends to be a more reliable indicator of
  convergence (see Figure~\ref{S-fig:convergence3d} of the Supporting Information).

  We reparameterized the time series, preserving the state sequence, in terms
  of the radial and axial coordinates ($r$, $z$) 
  %MRS: not needed.
  %because $x$ and $y$ aren't meaningful alone 
  due to the system's radial symmetry. We converted the $x$ and 
  $y$ center-of-mass coordinates to $r$. We fixed the state sequence from the 
  3D parameterization and applied the inference component of the IHMM procedure
  to the cylindrical trajectories in order to estimate the VAR(1) parameters in
  terms of $r$ and $z$. Since we fixed the state sequence, and we are using conjugate
  priors, the VAR parameters converge very quickly (see Figure~\ref{S-fig:fixed_state_convergence}).
  Therefore, we only ran the inference procedure for 100 iterations. 
  %MRS: hmm. Here the new covariance parameters is redone in r - does there need to be any transformation applied? 
  %MRS: though might be OK, since x^2 + y^2 = r^2 so maybe covariance are additive if x and y uncorrelated (which they should be)
  %However, note that the uncertainty in their
  %estimates are a function of the number of data points used in their estimate.

  We defined the final parameters of each state as the mean of the parameters 
  from each iteration recorded after the equilibration time point. We found the 
  equilibration time point of each component of the $\mathbf{c}$ vector as well as 
  $A$ and $\Sigma$ matrices, and then used the longest equilibration time of all 
  dimensions as the equilibration time point.

@@

  %MRS: maybe add subsubsections naming/labeling each part of the process so that people have a sense of the steps,
  %MRS: makes it easier to read.
  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size. For each solute studied, we identified 200-325
  independent states, each with separate VAR(1) parameters. Many of these states
  exhibit very similar dynamical behavior except their mean levels in $r$ and $z$
  are different.

  We reduced the parameter space used agglomerative clustering, a hierarchical
  clustering approach which uses a linkage criteria in order to successively merge
  similar clusters until a desired intracluster distance threshold or number of
  clusters is reached.~\cite{pedregosa_scikit-learn_2011} We used the Ward linkage 
  criteria, which works to minimize the sum of the squared differences within all
  clusters.~\cite{ward_hierarchical_1963} We elected to choose the number of clusters
  rather than the distance threshold. For our data, non-parametric methods such as 
  Bayesian Gaussian mixture models~\cite{pedregosa_scikit-learn_2011,gelman_bayesian_2013}
  tend to delocalize the clusters in parameter space (see Section~\ref{S-section:agglomerative}
  of the Supporting Information).
  %MRS: not clear what the consequence of the above is?

  We clustered based on the eigenvalues of $A$ and $\Sigma$, the radial means, $\mu_r$, 
  and the self-transition probabilities, $T_{ii}$, of each state for a total of 6 clustering
  dimensions. The means in $r$ are likely meaningful since the membrane is inhomogeneous, 
  transitioning from hydrophilic pores to hydrophobic tails. To cluster on $T_{ii}$, we first
  cast it in terms of the expected dwell times. The expected value of the number of sequential
  self-transitions is simply $\frac{1}{1 - T_{ii}}$. This relationship correctly implies that 
  dwell times approach infinity as $T_{ii}$ approaches 1. There we found the most success with
  agglomerative clustering by linearizing this relationship and clustering on $-\log(1 - T_{ii})$
  (see Section\ref{S-section:selfT_cluster} of the Supporting Information for a graphical
  illustration).

@@

  %BJC: to be reworked after I have a working procedure.
  Choosing the number of clusters can be somewhat subjective so
  we attempted to add some structure to the selection process by following
  a set of qualitative and quantitative guidelines. We used the silhouette 
  test in order to score the quality of clustering as a function of the number 
  of clusters chosen.~\cite{kaufman_finding_2009} For our data, the silhouette test generally favors the 
  lowest number of clusters possible (see Section~\ref{S-section:nclusters} of
  the Supporting Information). However, choosing too few clusters tends to 
  not distinguish between visually obvious differences in 
  dynamic behavior. This results in finalized parameter sets that are averages of
  distinct behavior which presents further problems with the predictive modeling
  that we discuss later on. We aimed to maintain the highest silhouette score,
  and thus lowest number of clusters, possible while verifying that visually 
  distinct states stayed separated. Based on these guidelines, we decided to 
  group the $A$ and $\Sigma$ parameters into five clusters each and the radial
  means into three clusters. Of the 75 distinct states allowed by this formulation,
  the solutes in this study showed behavior from 32--36 distinct clusters. While
  this may seem like a high number of states, many are sampled infrequently.
  %MRS: what is the consequence of being sampled infrequently.

@@

  %MRS: maybe ``using the parameterized model to explore mechanisms''
  \subsection{Tools for Exploring Mechanisms}\label{method:interactions}

@@

  %We developed a way to 
  We measure the local density of the membrane at arbitrary
  points in the unit cell as follows. We histogrammed the positions in three dimensions
  Since our system is in a monoclinic unit cell, we periodically replicated
  the system in the +/- x, y and z directions and then chose the bounds on the
  histogram in order to create a rectangular box encompassing the unit cell with
  a 1 nm buffer between the histogram and unit cell boundaries.~\cite{van_der_walt_numpy_2011} We then used a regular
  grid interpolator in order to allow interpolation at arbitrary points within the grid.~\cite{virtanen_scipy_2020}

@@

  Solutes in this system show a wide range of behavior influenced by the 
  heterogeneity of the membrane's nanostructure as well as the interactions 
  between monomer and solute chemical functionality. Our application of the 
  IHMM can identify and distinguish these behaviors. The following approach 
  to analysis demonstrates how one can discover and explain the very complex
  behavior exhibited by solutes over long MD trajectories in terms of 
  solute-membrane interactions. In addition, we can use the IHMM in order to generate
  stochastic trajectories, with dynamics similar to MD, so that we can estimate 
  macroscopic transport properties. 	
  %MRS: subsection lettering is really big - makes it seem like a very different section?

@@

  %MRS: maybe phrase it around types of mechanism, and use methanol as an example.
  \subsubsection*{The Dynamics of Methanol}

@@

  \begin{figure}
  \centering
  \begin{subfigure}{0.75\textwidth}
  \includegraphics[width=\textwidth]{mechanism_map.png}
  \caption{}
  \end{subfigure}
  \begin{subfigure}{0.24\textwidth}
  \includegraphics[width=\textwidth]{monomer_oxygens.png}
  \caption{}
  \end{subfigure}
  %MRS: a lot of this caption test is duplicated from the main text. I would suggest reducing the duplication.
  \caption{(a) We can learn a significant amount of detail about solute motion by viewing 
  solute trajectories color-coded according to their clustered dynamical behavior alongside
  plots of the dominant molecular interactions and descriptions of the membrane's local 
  environment. In the plot above, we show the radial and axial coordinates of a single
  methanol molecule over the course of the equilibrated portion of its MD simulation.
  Disjoint segments of the same color imply the solute behaves similarly at different 
  points of the trajectory because their VAR parameters place them in the same cluster. 
  There are eight distinct dynamical modes exhibited by this trajectory. In order
  of their first appearance, they are colored dark yellow, red, light blue, light green,
  orange, blue, teal and lime green. On the top of the plot are bars which represent different
  physical interactions and the membrane's local environment as a function of time. The top 
  bar is colored green at time points where solutes are associated with sodium. Methanol 
  associates with sodium ions relatively infrequently. The next bar down is colored when hydrogen
  bonds exist. The colors match the enlarged oxygen atoms highlighted in (b). Blue slivers
  correspond to hydrogen bonds donated to the monomer's carboxylate head group. Green slivers
  correspond to  hydrogen bonds donated to the ether linkages between the head groups
  and tails. Red slivers correspond to hydrogen bonds donated to the oxygens on the ends
  of the monomer tails. Methanol exhibits significant hydrogen bonding with all parts of
  the monomer. Which part of the monomer methanol hydrogen bonds to is heavily correlated
  to the solute's $r$ coordinate. Finally, the bar with the blue gradient represents the local 
  membrane density based on the solute's ($r$, $z$) coordinate. Darker shades of blue
  correspond to higher membrane densities. The density appears to be higher when methanol
  motion is restricted (e.g. from 75 to 550 ns). It is relatively low in areas where methanol's
  position fluctuates significantly (e.g. from 550 to 700 ns). 
  }\label{fig:mechanism_map}
  \end{figure}

@@

  We can begin to dissect the dynamics of each state by studying their parameters.
  States 1, 3 and 5 have radial means near 1 nm or less (see y axis of $r$ direction 
  plot in Figure~\ref{fig:common_states_MET_lines}), which suggest that they occur
  close to the pore centers. States 2, 4, 6 and 7 occur well within the tail region
  of the membrane. States 1 and 2 have higher covariances in both the axial and radial
  directions compared to all other states. Covariance in each dimension appears to have
  a strong correlation. States 3 and 5 (in the pore region) as well as states 4, 6 and
  7 (in the tail region) are primarily distinguished based on their autoregressive 
  coefficients, A. Higher values of A generally lead to more wandering about the state
  mean. One could argue that the parameters of states 3--7 are actually very close which
  suggests that solutes might behave alike in 
  %MRS: changed word.
  %topologically 
  morphologically
  distinct regions of the 
  membrane.

  Despite some similarity in their VAR parameters, the length of time spent in states 
  3--7 appears to be dependent on the solute's radial location. Based on the 
  self-transition probabilities, we estimated the average time spent by each solute in
  each state (Figure~\ref{fig:dwell_times_MET}). States 3, 5 and 7 have long dwell times
  relative to the rest because they have high self-transition probabilities. Somewhat 
  surprisingly, both states 3 and 5 correspond to solutes located close to a pore center,
  where one might expect solute motion to be less restricted. 
  %MRS: are thsse h-bonding states?
  Although solutes in states
  4 and 6 are in the viscous tail region, their dwell times are only intermediate. This
  could be due to inhomogeneity in the tail region which causes transitions between many
  different trapped states but with little net effect on solute MSD. However, there are
  instances of states within the tail region with dwell times that exceed 100--200 ns,
  as shown by our inclusion of state 7. There are no other states with dwell times longer
  than state 5 in the pore region.

@@

  The types of hydrogen bonding undergone by solutes is dependent on radial
  location (see Figure~\ref{fig:hbond_pichart}). States with radial means near 1 nm
  or less (states 1, 3 and 5) typically hydrogen bond with the monomer carboxylate
  groups and ether linkages. States with radial means greater than 1 nm (states 2,
  4, 6 and 7) tend to hydrogen bond with the monomer tails. 
  %MRS: tweak
  %In the past we've 
  Our previous work has
  shown that there is a gradual transition from the hydrophilic to hydrophobic region in
  these membranes, meaning the monomer location is not tightly bound in $r$. 
  %MRS: above logic is not clear?
  Therefore, there are instances of hydrogen bond interactions that one might consider 
  uncharacteristic based on the state's radial mean.
  %MRS: what is implication of ``uncharacteristicness''?

@@

  %MRS: maybe makes sense to talk a little about what features/parameters of the models the MSD is sensitive to?
  %BJC5: something we can do for the supporting information to show that this approach is valid: Eliminate nonidealities in our MD trajectories (compared to a pure VAR process) and use toy data to generate trajectories based on some number of states with different parameters. Generate an MSD with those. See if we can reproduce with our model. Can also use this to test new strategies.
  %MRS6: toy data is actually a VERY good idea.  Use a VAR process that looks like MD, and then try to cluster it, and figure out why different.

@@

  We showcase this modeling approach by example, but it is important to
  recognize the generality of this analysis, especially in the context of molecular
  simulations. Vector autogregressive models can describe a diverse set of behavior
  and so the IHMM may be suited to study many types motion, including those not 
  characterized by hopping and trapping. The IHMM is a powerful approach for 
  understanding particle motion as well as for generating inexpensive models
  which can give macroscopic insight.
